% !TeX root = ../incremental_SS_Translation.tex

\chapter{Uttaratantra 64: Preservation of Health}% svastharakṣanīya

\section{Literature}

Meulenbeld offered an annotated overview of this chapter, which is called 
\emph{svasthavṛtta} in the vulgate.\fvolcite{IA}[330--331]{meul-hist} There, 
he translates \emph{svasthavṛtta} as ``preservation of health'' but elsewhere 
as ``maintenance of health''\fvolcite{IA}[301]{meul-hist} and 
``hygiene.''\fvolcite{IB}[738]{meul-hist} His only reference to secondary 
literature on this topic is to an annotated German translation by J. Laping 
1984.\footnote{\cite{lapi-1984}, cited by \volcite{IB}[431]{meul-hist}.} His 
footnotes to the chapter also indicate 
many parallels with \emph{Carakasaṃhitā, Sūtrasthāna}, chapter six.

On the term \emph{svastha}, Meulenbeld cites W. Halbfass (1991): 250-252; M. 
Hara (1995) and P.V. Sharma (1985e: 52-53).\fvolcite{IB}[10]{meul-hist}

In primary sources, see the definition of \emph{svasthā} in 
\emph{Suśrutasaṃhitā, Sūtrasthāna} 15.41--44; \emph{svasthahita} in 
\emph{Aṣṭāṅgahṛdaya, Sūtrasthāna} 1.16cd; ...

Comparable chapters in other medical texts include the \emph{svasthacatuṣka} 
in \emph{Carakasaṃhitā, Sūtrasthāna} (see below) and a chapter on 
\emph{svasthavṛtti} in Cakrapāṇidatta's \emph{Cakradatta}.

Meulenbeld also refers to \emph{svasthacatuṣka}, a group of four chapters 
(5--8) in the \emph{Carakasaṃhitā, Sūtrasthāna}, chapters 5–8, also called 
\emph{svasthavṛttacatuṣka} or \emph{svāsthyavṛttikacatuṣka} in various 
colophons of the text and commentaries. He summarises these 
chapters,\fvolcite{IA}[13--14]{meul-hist} and their names are 
\emph{mātrāśitīya}, daily regime (cf. Bhela Sū.8; A.h.Sū.8; A.s.Sū.11. Cf. 
Su.Ci.24); \emph{tasyāśitīya}, seasonal regimen (\emph{ṛtucaryā}) 
(Meulenbeld cites many references in primary and secondary 
literature\fvolcite{IB}[13]{meul-hist}); \emph{navegāndhāraṇīya}, the 
unwholesome effects of suppressing natural urges, etc. (Cf. Bhela Sū.6; 
Su.Ci.24); and \emph{indriyopakramaṇīya}, senses, mind and rules of conduct 
(Cf. Bhela Sū.7; Su.Ci.24. See on Ca.Sū.8: S.Ch. Vidyabhusana (1971): 11--12).

Meulenbeld mentions that a ``Sadānanda wrote a commentary, called \emph{Auṣadhavivṛti}, on the chapters of the
\emph{Carakasaṃhitā} containing rules for the preservation of health 
(svasthavṛtta).\fvolcite{IA}[196]{meul-hist} His footnote on this work refers to 
the NCC (VI, 396–397) and the following bibliographic reference: 
\emph{pañcatantram \ldots\  
carakasūtrasthānasvasthavṛttacatuṣkākhyacaturadhyāyāḥ\ldots\
sadānandaśāstrikṛtauṣadhavivṛtiyutayā saṃvalitaṃ}, Mercantile Press, Lahore 
1926 [IO.San.D.554].\fvolcite{IB}[302]{meul-hist}

In the vulgate, this chapter is included in a section called, ``the chapters that 
are an embellishment of the text" (\emph{tantrabhūṣaṇādhyāya}). However, in 
the Nepalese version, it is the eighth part of tenfold section on treatment for the 
body (\emph{kāyacikitsita}). See K, folio 209r, ll. 4--5 
(\emph{mūtradoṣo mūtrāghāto yonyamānūṣa eva ca/  
apasmāronmādakañ caiva rasabhedas tathaiva ca// 
svastharakṣāvidhāṇañ ca tantrayuktiś ca doṣabhit/
ity ebhir daśakaṃ proktaṃ punaḥ kāyacikitsite//}).


\section{Translation}

\begin{translation}    
  
\item [1] 
Now we shall explain the upkeep of a healthy person. % rakṣaṇīya - maintenance of? prophylaxis for a healthy person?
	\footnote{The term \emph{rakṣaṇīya} (“upkeep”) is replaced in the vulgate \citep[808]{vulgate} by \emph{vṛtta}, which, when compounded with \emph{svastha}, denotes the “regime for a healthy person.” However, \emph{rakṣaṇa} is retained later in the same chapter of the vulgate.} 

\item [2] 
One whose humours, digestive fire, and the functioning of bodily constituents and impurities are in equilibrium; whose self, senses, and mind are serene is called healthy.\footnote{This verse is absent here in the vulgate but occurs at the end of \emph{Sūtrasthāna} 15 in both the Nepalese version and vulgate. The vulgate \citep[808]{vulgate} supplements its omission with the statement, ``Such a healthy person has been described in the \emph{Sūtrasthāna}'' (\emph{sūtrasthāne samuddiṣṭaḥ svastho bhavati yādṛśaḥ}). }

\item [3] 
O child, the upkeep of a healthy person is the supreme benefit of treatment. The topic of his regime and upkeep, which was mentioned by me at the beginning, was outlined in brief and will here be explained in detail.
% who are the interlocutors?
% something has gone wrong with artham in the Nepalese manuscripts (which has arthāṃ and arthān). The participle (uktaṃ) and verb (vakṣyate) require a singluar. And artham can be neuter. The syntax of the vulgate is wrong because the subject is plural (arthāḥ) and the verb singular (a patch made inadequate by metric causa?)
% ḍalhaṇa: āditaḥ anāgatabādhe

\item [4] 
In each season, when specific humours are aggravated in people, a wise physician should prescribe the appropriate tastes for that season.

\item [5] 
You should know that when people's digestive fire is weak because of the dampness of their bodies during the rainy months, wind and other humours become aggravated due to their excitement.
% Ḍalhaṇa appears to understand mārutādayaḥ as referring to the doṣas.
% have I over-translated khalu?

\item [6--8ab]
Therefore, in order to remove moisture and maintain the humours, the king should be given solid food with astringent, bitter, and pungent tastes, which is neither too oily nor excessively dry, and which is heating and stimulates digestive fire, along with the water which was mentioned before,%
        \footnote{Ḍalhaṇa\citep[809]{vulgate} understands “the water which was mentioned” as monsoon rains, which was described in the section on what to drink (\emph{yaj jalaṃ coktam ādita iti ‘varṣāsv antarikṣaṃ’ ityādinā pānīyavarge kathitam}).} %
or one should drink water that has been heated by hot iron and mixed with honey. 
% I am not sure about the yat in the fourth pāda

\item [8cd--9]
On a rainy and windy day that is affected by excessively cold water,  herbal medicines become acidic if they are fresh.%
	\footnote{Ḍalhaṇa \citep[809]{vulgate}  glosses \emph{vidāham} (“burning”) as \emph{amlapāka} (“sour digestion”), suggesting that fresh herbs may cause acidic irritation in the stomach when taken on a rainy, wind and cold wet day.}
	%
And because of that, a wise person should not undertake excessive physical exercise.

\item [10] % not in the vulgate
When it is not very cold, aged barley, wheat and rice with pungent broths and soups, as well as appetising wild meats, should be eaten. 
% emended to bhojyā based on Caraka Sū 6.38c (note in the edition)
% not sure of the standard translation of kaṭu as a taste (bitter, sharp, pungent?). 
% the ca in pāda 4 suggests that jāṅgala is wild meat/deer/game rather than an adjective. Elsewhere in Ḍalhaṇa I see eṇādayo jāṅgalamṛgamāṃseṣu (an adjective here but obviously refers to game).

\item [11] % not in the vulgate
One should drink a little mead or sweet wine, %
	\footnote{On \emph{mādhvīka}, a kind of honey liquour, see McHugh 2021: 46. On \emph{sīdhu} (“sweet wine”), Ḍalhaṇa \citep[778]{vulgate} says it is \emph{gauḍika}, an alcoholic drink produced from sugar or grain (\emph{gauḍikaṃ sīdhu guḍakṛtaṃ sasyakam}). McHugh (2021: 45–46), who translates \emph{sīdhu} as “sugarcane wine,” says that it is primarily made from sugarcane but also other sweet substances, like jaggery, herbs and mahua flowers.}
%
and also rainwater, lake water, or well water, hot or cold and mixed with honey. 

\item[12] % ab not in the vulgate
In this climate (\emph{atra}), one should avoid the easterly wind, exposure to the heat, buttered barley water, river water, sleep during the day, frost, and sex.

\item[13] % not in the vulgate
One who engages in rubbing, massage, bathing, perfumes, garlands and smoking in order to avoid the heat from the ground should sleep in the middle of the house. 
%
% uṣma heat, hot vapour, steam. Is humidity the right word here?
% cf: Caraka1.6.40: pragharṣodvartanasnānagandhamālyaparo bhavet | laghuśuddhāmbaraḥ sthānaṃ bhajed akledi vārṣikam || [during the rainy season]
% cf: AS.Sū.4.48	pragharṣodvartanasnānadhūmagandhāgarupriyaḥ | 
%				yāyāt kareṇumukhyābhiścitrasragvastrabhūṣitaḥ || [varṣa season]
% Jayaratha on Tantrāloka: madhyegṛhamiti--gṛhamadhye brahmasthāne--ity arthaḥ
% madhyamaveśmani might denote 'in an inner/central chamber of the house.'




\end{translation}    
